---
title: "Assignment 1"
author: "Emanuele Coradin"
date: "2024-04-09"
output: 
  read_document: rmdformats::readthedown
  html_document:
    number_sections: true
    theme: spacelab
  pdf_document:
    number_sections: true
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

color_vector <- c("#CC0000",   # Dark red
                  "#DB1E60",   # Pinkish-red
                  "#CC79A7",   # Muted purple
                  "#D55E00",   # Vermilion
                  "#009E73",   # Bluish green
                  "#56B4E9",   # Sky blue
                  '#000046',   # Deep Blue
                  "#E69F00")   # Yellow-orange
```

# Exercise 1

The repository <https://drive.google.com/drive/folders/1NESuaJ5yGIrAli1TgrpnK5hnoxGsMi3h?usp=sharing> contains bike-sharing data provided by New York City, Citi Bike sharing system. The data (in csv format) is structured as follows:

-   Trip duration (in seconds)
-   Start Time and date
-   Stop Time and date
-   Start Station ID, name, latitude and longitude
-   End Station ID, name, latitude and longitude
-   Bike ID
-   User Type (Customer or Subscriber )
-   Birth’s Year
-   Gender (0=unknown; 1=male; 2=female)

## Points 1-3

1.  Read the data and import them in a data.frame or tibble structure;
2.  Merge the five data frames in a unique structure;
3.  Check for missing data and remove it.

The files are stored inside the \*\*Data_CitiBike\*\> **folder, the named** JC-20190x-citibike-tripdata.csv\*\* with x in {2, 3, 4, 5, 6} being the identifier of the file.

```{r, message=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
library(geosphere)
```

```{r}
# ---------------- READING AND IMPORTING DATAFRAMES ----------------
PATH <- "./Data_CitiBike/"
indices <- 2:6
files <- unlist(lapply(indices, function(i) {
  sprintf("%sJC-20190%i-citibike-tripdata.csv", PATH, i)
}))

dataframes <- lapply(files, function(filename) {
  read.csv(filename, header = TRUE, sep = ",")
})

# Merging
df <- bind_rows(dataframes)

# Checking for missing data
#fill blanks with not a numbers
df[df==""]<-NA
#complete.cases() gives TRUE if the row contains all the fields
df<-df[complete.cases(df),]

head(df)

```

## Point 4

4.1 Compute the average and the median trip duration in minutes

4.2 Evaluate the minimum and maximum trip duration; does that sound like a reasonable value?

4.3 Repeat the calculation of the average (and the median) trip duration by excluding trips longer than 3 hours. Next, evaluate the number of skimmed entries 
4.4 Plot the distribution of trip duration after the skimming of the previous point

```{r }
writeLines(sprintf("Average trip duration:   %.1f min 
                  \nMedian trip duration:    %.1f min 
                  \nMaximum trip duration:   %.1f min 
                  \nMinimum trip duration:   %.1f min", 
               mean(df$tripduration)/60.,
               median(df$tripduration)/60.,
               max(df$tripduration)/60.,
               min(df$tripduration)/60.))
```

```{r }
skimmed_df = df[df$tripduration<(3.*3600), ]
writeLines(sprintf(
              "Average trip duration:   %.1f min
             \nMedian trip duration:    %.1f min
             \nMaximum trip duration:   %.1f min
             \nMinimum trip duration:   %.1f min
             \nNumber of rejected rows: %i", 
               mean(skimmed_df$tripduration)/60.,
               median(skimmed_df$tripduration)/60.,
               max(skimmed_df$tripduration)/60.,
               min(skimmed_df$tripduration)/60.,
               nrow(df) - nrow(skimmed_df)))

```

```{r }
p<-ggplot(skimmed_df, aes(tripduration/60, y = after_stat(density))) + 
  geom_histogram(color="black", fill=color_vector[6]) +
  geom_density(alpha=.3, fill=color_vector[2], linetype="dashed") +
  labs(
    title = "Trip duration distribution",
    x = "Duration [min]",
    y = "PDF [1/min]"
  ) +
  theme(plot.title = element_text(hjust = 0.5))+
  scale_x_continuous(trans = "log", breaks = c(1, 10, 100))

print(p)
```

## Point 5

Plot the monthly average trip duration

```{r }
month <- unlist(lapply(skimmed_df$starttime, function(date) {
  as.POSIXlt(date)$mon
}))

skimmed_df$month = month
```

```{r }
monthly_trip <- skimmed_df %>%
  group_by(month) %>%
  summarise(average_duration = mean(tripduration) / 60, std = sd(tripduration)/60)

p<- ggplot(monthly_trip, aes(x = month, y = average_duration)) + 
  geom_line(color=color_vector[1]) +
  geom_point(color=color_vector[7])+
  geom_errorbar(aes(ymin=average_duration-std, ymax=average_duration+std), width=.2, color = color_vector[7], alpha = 0.7,
  position=position_dodge(0.05)) +
  labs(
    title = "Average Trip duration per Month",
    x = "Month",
    y = "Duration [min]"
  )+
  theme(plot.title = element_text(hjust = 0.5))
print(p)

```

## Point 6

6.1 Plot the number of rides per day

6.2 Plot the hourly distribution on weekdays and weekends

6.3 Plot the hourly distribution on weekdays and weekends separating by customers and users

```{r }
date <- unlist(lapply(skimmed_df$starttime, function(date) {
  as.POSIXct(date, format = "%Y-%m-%d")
}))

skimmed_df$date = date

hour <- as.numeric(format(as.POSIXct(skimmed_df$starttime), format = "%H"))
skimmed_df$hour <- hour

day_of_week <- weekdays(as.POSIXct(skimmed_df$starttime))
skimmed_df$day_of_week <- day_of_week

weekends_list = c('sabato', 'domenica')
weekends <- skimmed_df$day_of_week %in% weekends_list
skimmed_df$weekends = weekends

```

```{r }
rides_per_day <- skimmed_df %>%
                 group_by(date) %>%
                 summarise(rides = n())

p <- ggplot(rides_per_day, aes(x = as.POSIXct(date), y = rides)) + 
  geom_line(color = color_vector[1])  +
  geom_point(color = color_vector[7]) +
  labs(
    title = "Number of Rides per Day",
    x = "Date",
    y = "Number of Rides"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

print(p)

```

```{r }
hourly_distro <- skimmed_df %>%
                 group_by(weekends, hour) %>%
                 summarise(rides = n())

p <- ggplot(hourly_distro, aes(x = hour, y = rides, color = factor(weekends))) + 
  geom_line()  +
  geom_point(aes(color = factor(weekends))) +
  
  labs(
    title = "Number of Rides per Hour",
    x = "Hour",
    y = "Number of Rides",
    color = "Weekend"
  ) +
  scale_color_manual(values = c(color_vector[1], color_vector[7])) +
  scale_y_continuous(trans = "log", breaks = c(10, 100, 1000, 10000, 100000)) +
    
  theme(plot.title = element_text(hjust = 0.5))

print(p)

```

```{r }
hourly_distro <- skimmed_df %>%
                 group_by(weekends, hour, usertype) %>%
                 summarise(rides = n())

p <- ggplot(hourly_distro, aes(x = hour, y = rides, color = factor(weekends), linetype = factor(usertype))) + 
  geom_line()  +
  geom_point(aes(color = factor(weekends))) +
  
  labs(
    title = "Number of Rides per Hour",
    x = "Hour",
    y = "Number of Rides",
    color = "Weekend",
    linetype = "User role"
  ) +
  scale_color_manual(values = c(color_vector[1], color_vector[7])) +
  scale_y_continuous(trans = "log", breaks = c(10, 100, 1000, 10000, 100000)) +
  theme(plot.title = element_text(hjust = 0.5))

print(p)
```

## Point 7

7.1 Evaluate the average speed of users

7.2 Plot the average speed as a function of route length

7.3 Separate the results for weekdays and weekends

```{r }
distances <- unlist(
  mapply(
    function(lon1, lat1, lon2, lat2){
      distHaversine(c(lon1, lat1), c(lon2, lat2))
    },
    skimmed_df$start.station.longitude, skimmed_df$start.station.latitude, skimmed_df$end.station.longitude, skimmed_df$end.station.latitude
  ))
skimmed_df$distances <- distances
average_speed <- skimmed_df$distances / skimmed_df$tripduration
skimmed_df$average_speed = average_speed
writeLines(sprintf(
              "Average speed:   %.1f m/s",
              mean(average_speed)
              ))
```

```{r }
p <- ggplot(skimmed_df[skimmed_df$average_speed*3.6<30 & skimmed_df$average_speed>0 & skimmed_df$distances < 7.5e03, ], aes(x = distances/1000, y = average_speed*3.6)) + 
  geom_bin2d(bins = 100) +
  scale_fill_continuous(type = "viridis") +
  labs(
    title = "Average Speed vs route length",
    x = "Length [km]",
    y = "Average Speed [km/h]"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

print(p)
```

```{r }
facet_labels <- c(
  "TRUE"  = "Weekends",
  "FALSE" = "Weekdays"
)
p <- ggplot(skimmed_df[skimmed_df$average_speed*3.6<30 & skimmed_df$average_speed>0 & skimmed_df$distances < 7.5e03, ], aes(x = distances/1000, y = average_speed*3.6)) + 
  geom_bin2d(bins = 100, aes(fill = after_stat(density))) +
  scale_fill_continuous(type = "viridis", name="Density") +
  labs(
    title = "Average Speed vs route length",
    x = "Length [km]",
    y = "Average Speed [km/h]"
  ) +
  facet_wrap(~ weekends, labeller = as_labeller(facet_labels)) +
  theme(plot.title = element_text(hjust = 0.5))

print(p)
```

## Point 8

8.1 Find the most common start station and the least popular end station.

8.2 Show the distribution of start stations

8.3 Find the three most common routes and the three least popular ones

```{r }
start_station_count <- skimmed_df %>% 
                       group_by(start.station.id) %>%
                       summarise(start_count = n()) %>%
                       arrange(desc(start_count))

end_station_count   <- skimmed_df %>% 
                       group_by(end.station.id) %>%
                       summarise(end_count = n()) %>%
                       arrange(end_count)

route_count         <- skimmed_df %>% 
                       group_by(start.station.id, end.station.id) %>%
                       summarise(route_count = n()) %>%
                       arrange(desc(route_count))
```

```{r }
n_routes = nrow(route_count)
writeLines(
  sprintf(
"
The most common start station has ID: %i 
The least popular end station has ID: %i
The three most common routes are (start_id, end_id):
  - (%i,%i);
  - (%i,%i);
  - (%i,%i).
The three least common routes are (start_id, end_id):
  - (%i,%i);
  - (%i,%i);
  - (%i,%i).
"   , 
    start_station_count[[1,1]],
    end_station_count[[1, 1]],

    route_count[[1, 1]], route_count[[1, 2]], 
    route_count[[2, 1]], route_count[[2, 2]],
    route_count[[3, 1]], route_count[[3, 2]],

    route_count[[n_routes, 1]],   route_count[[n_routes, 2]],
    route_count[[n_routes-1, 1]], route_count[[n_routes-1, 2]],
    route_count[[n_routes-2, 1]], route_count[[n_routes-2, 2]]
  )
)
```
```{r }
p<-ggplot(start_station_count, aes(x = as.character(start.station.id), y = start_count)) +
  geom_bar(stat = "identity", fill = color_vector[6], color = "black") +
  labs(
    title = "Distribution of start stations",
    x = "Start station ID",
    y = "Counts"
  ) +
  theme(plot.title = element_text(hjust = 0.5),
  axis.text.x = element_text(angle = 60, hjust = 1))

print(p)
```
