---
title: "Models Inference"
author: "Emanuele Coradin"
date: "2024-04-08"
output: 
  read_document: rmdformats::readthedown
  html_document:
    number_sections: true
    theme: spacelab
  pdf_document:
    number_sections: true
    toc: true
    toc_depth: 2
---

------------------------------------------------------------------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Model comparison

Given a set of model we want to find out which one is best supported by data, computing and comparing the Posterior Probability

## Bayesian Model Comparison

In the simplest case we compare models with no parameters.

### Problem

- A test for a disease is 90% reliable.
- The probability of testing positive, in the absence of the disease, is 0.07.
- Among people aged 40 to 50 with no symptoms, 8 in 1000 have the disease.

If a person in his/her 40 tests positive, what is the probability that he/she has the
disease ?

We build the following propositions:
- D: a person is tested positive
- H: a person has the disease

So the probabilities are: 
- P(D|H) = 0.9
- P(D|H* = 0.07
- P(H) = 0.008

```{r}

post <- function(p.d.m, p.d.notm , p.m) {
  p.notm <- 1 - p.m
  odds.ratio <- (p.d.m * p.m) /
  (p.d.notm * p.notm)
  p.m.d <- 1/(1 + 1/odds.ratio)
}
p.d.m <- seq(0, 1, 0.01) # True positive
p.d.notm <- 0.07 # False positive
p.m <- 0.008 # Disease Prior
p.m.d <- post(p.d.m, p.d.notm , p.m)
plot(p.d.m, p.m.d, type='l', lwd=2, col='navy')
p.d.m <- 0.9 # True positive
p.d.notm <- 10^seq(-4,0, 0.02) # False positive
p.m <- 0.008 # Disease Prior
p.m.d <- post(p.d.m, p.d.notm , p.m)
plot(log10(p.d.notm), p.m.d, type='l', col='navy')

```

## Occam's Razor


## Generative models

Theory predicting observable data from model parameters.
In real life data is always affected by noise: we can also construct noise models "*Measurement model*".





